### 2018-04-07 2_wmm.tex
Complete the data-flow encoding.

%"Semantics and verification under weak memory models have been the subject of study at least since 2007." [PORTHOS]
The \textit{weak memory model} is a set of predicative constraints on possible executions of a concurrent program. The study of formalisation weak memory models for different architectures has been rapidly developed over last decade.

Firstly, the research of WMM aimes to formalise the weak memory models and provide systematic, sound and complete formal approach of defining WMMs in order to be able to verify systems with respect to them.
A generic framework of program analysis with respect to weak memory models was presented in 2010~\cite{alglave2010shared}. Similar approach is used \texttt{PORTHOS} as it is described
% in Section~\cite{ch:wmm:event}. 
further in current Chapter.

@@

The event-based program model represents the directed graph (\textit{event-graph}), where vertices represent \textit{events}, and edges represent \textit{relations} over the events. An event is something, that, after being executed, changes the state of an abstract machine executing the concurrent program. 
An \textit{execution} (trace, run) of a given program is an ordered set of events. The order of events in particular execution is denoted as `$\rightarrow$', an empty execution is denoted as $\emptyset$.
%associated with the instructions of the program
%the path in the event-graph. 
An execution is considered to be \textit{valid} if the memory events follow a single global timeline, textit.e., can be embedded in a single partial order allowed by the memory model restrictions~\cite{alglave2010shared}. An execution to be checked on validity is called the \textit{candidate execution}.

@@

%The \textit{control-flow} instructions (conditional and unconditional jumps) are encoded into the model directly, without additional events, as the $po$-relation (for \textit{program order}; see Chapter~\ref{ch:wmm:model:relations} for detailed definition of relations).
The \textit{control-flow} of a program is defined by \texttt{po}-relation $\subset~\mathbb{E}~\times~\mathbb{E}$ (\textit{program-order}), which represents the total order of events \textit{from same process}, which never relates events from different processes.
Thus, if a program specifies the memory instruction $i_2$ to follow immideately the memory instruction $i_2$, then there exist an edge $e_1 \xrightarrow{\mathtt{po}} e_2$ in the event-graph where event $e_1$ is caused by the instruction $i_1$ and $e_2$ is caused by the instruction $i_2$. This relation encodes the control-flow of the program into the event-graph.
%Some new relations may be acquired : dp, po-loc

The \textit{data-flow} of a program is defined by \textit{communication relations}: the \texttt{rf}-relation $\subset~\mathbb{W}~\times~\mathbb{R}$ (\textit{read-from} relation) that maps a write to a read reading its value, the \texttt{co}-relation $\subset~\mathbb{W}~\times~\mathbb{W}$ (\textit{coherence order}, sometimes called \texttt{ws}-relation for \textit{write serialisation}) defines the total order on writes to the same location across all processes, and the \texttt{fr}-relation $\subset~\mathbb{R}~\times~\mathbb{W}$ (\textit{from-read order}) that maps a read to possible writes preceding the current write event (this relation is the inversion of the \texttt{rf}-relation: $\mathtt{rf} = \mathtt{fr}^{-1}$).
%TODO: perephrase last sentence! alglave thesis, p. 36

Figure~\ref{simple_wmm_x86_pic} illustrates the candidate execution for the Example~\ref{simple_wmm_x86}, that reaches the state \texttt{(0:EAX=0~/\textbackslash~1:EAX=0)} within x86-TSO memory model as it forbids cycles over \texttt{po}$\cup$\texttt{rf} (union) relation (the picture is generated by the \texttt{herd7} tool, version 7.47). 
%The event $b$:\texttt{(Ry=0)} reads value \texttt{0} at the shared location \texttt{y} from the initial write event $e$\texttt{(Wy=0)} (the red edge of \texttt{rf}-relation), consequently, 
%the event $d$:\texttt{(Rx=0)} reads value \texttt{0} at the shared location \texttt{x} from the initial write event $f$\texttt{(Wx=0)} (the red edge of \texttt{rf}-relation). 
%However, the Power memory model allows such cycles, therefore %TODO: include load buffering example, see presentation pdf

@@

//to be done

>>>>

%"Semantics and verification under weak memory models have been the subject of study at least since 2007." [PORTHOS]
A \textit{weak memory model} is a set of predicative constraints over so-called \textit{derived relations}, which defines the set of possible executions of a concurrent program.
Usually, these constraints include requirements of the reflexivity or acyclicity of derived relations.
The study of formalisation weak memory models for different architectures has been rapidly developed over last decade.

Firstly, the research of WMM aimes to formalise the weak memory models and provide systematic, sound and complete formal approach of defining WMMs in order to be able to verify systems with respect to them.
A generic framework of program analysis with respect to weak memory models was presented in 2010~\cite{alglave2010shared}. Similar approach is used \Porthos as it is described
% in Section~\cite{ch:wmm:event}. 
further in current Chapter.

@@

The event-based program model represents the directed graph (the \textit{event-flow graph}), where vertices represent \textit{events}, and edges represent \textit{relations} over the events. An event is something, that, after being executed, changes the state of an abstract machine executing the concurrent program. 
An \textit{execution} (trace, run) of a given program is an ordered set of events. The order of events in particular execution is denoted as `$\rightarrow$', an empty execution is denoted as $\emptyset$.
%associated with the instructions of the program
%the path in the event-flow graph. 
An execution is considered to be \textit{valid} if the memory events follow a single global timeline, textit.e., can be embedded in a single partial order allowed by the memory model restrictions~\cite{alglave2010shared}. An execution to be checked on validity is called the \textit{candidate execution}.

@@

In this section, we describe basic derived relations used in memory model-awared program analysis.

%The \textit{control-flow} instructions (conditional and unconditional jumps) are encoded into the model directly, without additional events, as the $po$-relation (for \textit{program order}; see Chapter~\ref{ch:wmm:model:relations} for detailed definition of relations).
The \textit{control-flow} of a program is defined by \texttt{po}-relation $\subset~\mathbb{E}~\times~\mathbb{E}$ (\textit{program-order}), which represents the total order of events \textit{from same process}, which never relates events from different processes.
Thus, if a program specifies the memory instruction $i_2$ to follow immideately the memory instruction $i_2$, then there exist an edge $e_1 \xrightarrow{\mathtt{po}} e_2$ in the event-flow graph where event $e_1$ is caused by the instruction $i_1$ and $e_2$ is caused by the instruction $i_2$. This relation encodes the control-flow of the program into the event-flow graph.
%Some new relations may be acquired : dp, po-loc

The \textit{data-flow} of a program is defined by \textit{communication relations}: the \rf-relation $\subset~\mathbb{W}~\times~\mathbb{R}$ (\textit{read-from} relation) that maps a write to a read reading its value, the \co-relation $\subset~\mathbb{W}~\times~\mathbb{W}$ (\textit{coherence order}, sometimes called \texttt{ws}-relation for \textit{write serialisation}) defines the total order on writes to the same location across all processes, and the \fr-relation $\subset~\mathbb{R}~\times~\mathbb{W}$ (\textit{from-read order}) that maps a read to possible writes preceding the current write event (this relation is the inversion of the \rf-relation: $\mathtt{rf} = \mathtt{fr}^{-1}$).
%TODO: perephrase last sentence! alglave thesis, p. 36

Figure~\ref{simple_wmm_x86_pic} illustrates the candidate execution for the Example~\ref{simple_wmm_x86}, that reaches the state \texttt{(0:EAX=0~/\textbackslash~1:EAX=0)} within x86-TSO memory model as it forbids cycles over \po$\cup$\rf (union) relation (the picture is generated by the \texttt{herd7} tool, version 7.47). 
%The event $b$:\texttt{(Ry=0)} reads value \texttt{0} at the shared location \texttt{y} from the initial write event $e$\texttt{(Wy=0)} (the red edge of \texttt{rf}-relation), consequently, 
%the event $d$:\texttt{(Rx=0)} reads value \texttt{0} at the shared location \texttt{x} from the initial write event $f$\texttt{(Wx=0)} (the red edge of \texttt{rf}-relation). 
%However, the Power memory model allows such cycles, therefore %TODO: include load buffering example, see presentation pdf

@@

//axioms of TSO wmm

// example of sets: rf, co, ... for the code snipped used before

### 2018-04-07 3_encod.tex
Complete the data-flow encoding.

\subsection{Encoding the control-flow constraints}
\label{ch:port:enc:cf}

@@

While working on the \tool{mousquetaires}, we applied some optimisations of the encoding scheme for the control-flow in order to decrease the size of result formula. 
The \tool{PORTHOS} tool encodes the \textit{instructions}, which have recursive nature, in a recursive manner. This means, it adds synthetic composite instructions for both linear (sequential) and non-linear (branching) instructions. 
For instance, control-flow of the sequential instruction \m{i_1 := i_2; i_3} is encoded as 
\m{\phi_{CF}(i_2;i_3) = (cf_{i_1} \Leftrightarrow (cf_{i_2} \land cf_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)}, 
and control-flow of the branching instruction \m{i_1 := \{c \, ? \, i_2 \, : \, i_3\}} is encoded as 
\m{\phi_{CF}( \, c \, ? \, i_2 \, : \, i_3) = (cf_{i_1} \Leftrightarrow (cf_{i_2} \lor cf_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)} (here we used the notation of C-like ternary operator `\texttt{x?y:z}' for defining the conditional expression `\texttt{if} $x$ \texttt{then} $y$ \texttt{else} z').

In contrast, the \tool{mousquetaires} firstly compiles the recursive high-level code into the linear low-level event-based representation,
% (similar to an assembly language)
that is then encoded into SMT formula. The encoding of branching nodes depends on the \textit{guards}, the value of conditional variable on the branching state, which in turn is encoded as data-flow constraint (see further in current Chapter).
%linearly as it shown in

@@

For listed cases, below we propose the encoding scheme that uniquely encodes each node of graph and allows to encode partially executed program.
The Equation~\ref{enc:cf:seq} for encoding the sequential control-flow represented on Figure~\ref{encode:cf:seq} reflects the fact that the event $e_2$ can be executed iff the event $e_1$ was executed. The Equation~\ref{enc:cf:br} for encoding the branching control-flow depicted on Figure~\ref{encode:cf:br} allows only following executions: $\{\emptyset, (e_1), (e_1 \rightarrow e_2), (e_1 \rightarrow e3) \}$. In encoding~\ref{enc:cf:merge} of the merge-point represented on Figure~\ref{encode:cf:merge}, the event $e_k$ is executed if either of its predecessors was executed, regardless of type of the transition.

\begin{align}
%\phi_{CF_{(e_1;e_2)}}     =
%\phi_{CF_{(e_1?e_2:e_3)}} =
\phi_{CF_{seq}}   = \ & \fx(e_2) \rightarrow \fx(e_1) \label{enc:cf:seq} \\
\phi_{CF_{br}}    = \ & [\fx(e_2) \rightarrow \fx(e_1)] \land [\fx(e_2) \rightarrow \fv(e_1)] \nonumber \\
				  & \land  [\fx(e_3) \rightarrow \fx(e_1)] \land [\fx(e_3) \rightarrow \lnot\fv(e_1)] \nonumber \\
				  & \land \lnot [\fx(e_2) \land \fx(e_3)]  \label{enc:cf:br} \\
\phi_{CF_{merge}} = \ & \fx(e_k) \rightarrow (\bigvee\limits_{e_p \in \mathtt{pred}(e_k)}^{} \fx(e_p)) \label{enc:cf:merge}
\end{align}

\subsection{Encoding the data-flow constraints}
\label{ch:port:enc:df}

While encoding the data-flow constraints, we use the \textit{Static Single-Assignment (SSA) form} in order to be able to capture an arbitrary data-flow into a single SMT-formula. The single-assignment requires that a variable is not assigned once it has been accessed (either read or written). That is, on each state of the program a variable (either local or shared) is mapped to its indexed reference, which represents a new boolean variable in the formula. On merge-points, the $\phi$-function is emulated as the disjunction of values of the variable, that is computed as the reaching definition. For that, we keep track of the map that for each event stores mapping from all declared variables to their SSA-indexes. All indexes are globally unique within single program analysis.

As it is implemented in~\cite{Porthos17}, the SSA-indexes of variables are computed in accordance with the following rules: (1) any access to a shared variable (both read and write) increments its SSA-index; (2) only writes to a local variable increment its SSA-index (reads preserve indexes); (3) no access to a constant variable or computed expression changes their SSA-index.

\subsection{Encoding the memory model constraints}
\label{ch:port:enc:wmm}

>>>>

\subsection{The control-flow encoding} %Encoding the control-flow constraints}
\label{ch:port:enc:cf}

@@

While working on the \tool{mousquetaires}, we applied some modifications of the encoding scheme for the control-flow. The changes are conditioned by the need to be able to process an arbitrary control-flow produced by conditional and unconditional jumps of C language. 
For that, we compile the recursive abstract syntax tree~(AST) of the parsed C-code to the plain (non-recursive) event-flow graph.
We show%TODO! to show, actually
that the new encoding is smaller than the old one used in \Porthos since it does not produces new variables for each high-level statement of the input language.
%The \tool{PORTHOS} tool encodes the \textit{instructions}, which have recursive nature, in a recursive manner. This means, it adds synthetic composite instructions for both linear (sequential) and non-linear (branching) instructions. 
For instance, in old scheme the control-flow of the sequential instruction \m{i_1 := i_2; i_3} was encoded as 
\m{\phi_{CF}(i_2;i_3) = (cf_{i_1} \Leftrightarrow (cf_{i_2} \land cf_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)}, 
and control-flow of the branching instruction \m{i_1 := \{c \, ? \, i_2 \, : \, i_3\}} was encoded as 
\m{\phi_{CF}( \, c \, ? \, i_2 \, : \, i_3) = (cf_{i_1} \Leftrightarrow (cf_{i_2} \lor cf_{i_3})) \land \phi_{CF}(i_2) \land \phi_{CF}(i_3)} (here we used the notation of C-like ternary operator `\texttt{x?y:z}' for defining the conditional expression `\texttt{if} $x$ \texttt{then} $y$ \texttt{else} z').
In contrast, the new scheme implemented in \tool{mousquetaires} firstly compiles the recursive high-level code into the linear low-level event-based representation, that is then encoded into an SMT-formula. The encoding of branching nodes depends on the \textit{guards}, the value of conditional variable on the branching state, which in turn is encoded as data-flow constraint (see further in current Chapter).
%linearly as it shown in

@@

For listed cases, below we propose the encoding scheme that uniquely encodes each node of graph and allows to encode partially executed program.
The Equation~\ref{enc:cf:seq} for encoding the sequential control-flow represented on Figure~\ref{encode:cf:seq} reflects the fact that the event $e_2$ can be executed iff the event $e_1$ was executed. The Equation~\ref{enc:cf:br} for encoding the branching control-flow depicted on Figure~\ref{encode:cf:br} allows only following executions: $\{\emptyset, (e_1), (e_1 \rightarrow e_2), (e_1 \rightarrow e3) \}$. In encoding~\ref{enc:cf:merge} of the merge-point represented on Figure~\ref{encode:cf:merge}, the event $e_k$ is executed if either of its predecessors was executed, regardless of type of the transition.
%TODO: maybe add the overall formula \phi_{CF} ?

\begin{align}
%\phi_{CF_{(e_1;e_2)}}     =
%\phi_{CF_{(e_1?e_2:e_3)}} =
\phi_{CF_{seq}}   = \ & \fx(e_2) \rightarrow \fx(e_1) \label{enc:cf:seq} \\
\phi_{CF_{br}}    = \ & [\fx(e_2) \rightarrow \fx(e_1)] \land [\fx(e_3) \rightarrow \fx(e_1)] \land \nonumber \\
				  &  [\fx(e_2) \rightarrow \fv(e_1)] \land [\fx(e_3) \rightarrow \lnot\fv(e_1)] \land \nonumber \\
				  & \lnot [\fx(e_2) \land \fx(e_3)]  \label{enc:cf:br} \\
\phi_{CF_{merge}} = \ & \fx(e_k) \rightarrow (\bigvee\limits_{e_p \in \mathtt{pred}(e_k)}^{} \fx(e_p)) \label{enc:cf:merge}
\end{align}

\subsection{The data-flow encoding} %constraints}
\label{ch:port:enc:df}

While encoding the data-flow constraints, we use the \textit{Static Single-Assignment~(SSA) form} in order to be able to capture an arbitrary data-flow into a single SMT-formula. The SSA form requires each variable to be assigned only once within entire program. That is, on each state of the program a variable (either local or shared) is mapped to its indexed reference, which represents a new boolean variable in the formula. On merge-points, the $\phi$-function is emulated as the disjunction of values of the variable that is computed as the \textit{reaching definition}. For that, we carry the map that for each event stores mapping from all declared variables to their SSA-indices. All indices are globally unique within single program analysis.

As it is implemented in~\cite{Porthos17}, the SSA-indices of variables are computed in accordance with the following rules: (1) any access to a shared variable (both read and write) increments its SSA-index; (2) only writes to a local variable increment its SSA-index (reads preserve indices); (3) no access to a constant variable or computed expression changes their SSA-index.
These rules determine the following encoding of load, store and computation events within single thread:
%TODO: maybe add the overall formula \phi_{DF} ?

\begin{align}
\phi_{DF_{e = \mathtt{load}(r \leftarrow l)}} = \ & \fx(e) \rightarrow (r_{i+1} = l_{i+1}) \\
\phi_{DF_{e = \mathtt{store}(l \leftarrow r)}} = \ & \fx(e) \rightarrow (l_{i+1} = r_i) \\
\phi_{DF_{e = \mathtt{compute}(...)}} = \ & \fx(e) \rightarrow \fv(e) % \\
\end{align}

Comparing to the old \Porthos version, the algorithm of converting the program into the SSA form was also adapted for processing the plain flow-graph data structure. That is, for each node we keep track the set of its predecessors, and then, while traversing the graph in topological order, we propagate the SSA-indices for each variable defined so far, and update these indices according the rules listed above.

The time of described algorithm is linear of the size of event-flow graph since it performs only single traverse of the graph. Notwithstanding the overhead of storing (or, equivalently but undesirably, computing with%TODO: are you sure that 'with'?
linear time) the predecessors of each node in order to be able to convert the program into an SSA form, the time of such a tranformation reduced%TODO: How much ????
comparing to the algorithm implemented in \Porthos, which recomputed SSA-indices recursively for each instruction.%TODO: PROOF OR REMOVE !  Maybe, some words of copying, merging SSA maps ?

Moreover, the data-flow encoding scheme used in \Porthos faced the problem when the branching statements had different number of assignments of certain variable in their branches. //TODO: why? why don't we have this problem now?

As it has been stated before, the \rf-relation links the parts%TODO: another word?
of data-flow stored in equivalence assertions over the SSA-variables. 
%The encoding of the \rf-relation, which links the SSA-variables to the original variables, 
The encoding of this linkage left untouch as it is implemented in \Porthos: for each pair of events $e_1$ and $e_2$ linked by the \rf-relation, we add the following constraint:

\begin{align}
\phi_{DF_{mem}}(e_1, e_2) = \ & \mathtt{rf}(e_1, e_2) \rightarrow (l_i = l_j) 
\end{align}

where the variable of location $l$ is mapped to the SSA-variable $l_i$ for event $e_1$, and to the SSA-variable $l_j$ for event $e_2$; and the predicate $\mathtt{rf}(e_1, e_2)$ is encoded as a boolean variable, which itself equals $true$ if $e_2$ reads the shared variable that was written in $e_1$.

\subsection{The memory model encoding} %Encoding the memory model constraints}
\label{ch:port:enc:wmm}

The encoding of a weak memory model itself does not depend on the analysing program. The problem here...

recursive definitions and Kleene iteration

As it is described in~\cite{Porthos17}, the is based ...
%~\cite{Porthos17}%, Section "Programs and Memory Models", subsection "Encoding Derived Relations"
